---
title: "Apple firmness — vibration-trace processing pipeline"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: show
    fig_width: 7
    fig_height: 4
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center")
```

## Background

Portfolio piece — a synthetic recreation of a non-destructive
**vibration-based firmness** measurement workflow I built during a
fruit-quality project. The original recordings are not shareable, so this
repo simulates qualitatively similar two-channel waveforms (stem end
and calyx end), recovers the characteristic minimum from each trace,
extracts its time `t_min`, and looks at the joint
`(CH1 t_min, CH2 t_min)` distribution.

The pipeline reproduces the four steps I used in practice:

1. acquire two-channel time-series per fruit;
2. detect the diagnostic minimum on each channel;
3. plot every trace with a coloured marker at the detected minimum
   for visual QA;
4. summarise the dataset as a `CH1` × `CH2` scatter coloured by trial.

## Simulated data

```{r simulate, results = "hide"}
source("R/simulate.R")
```

```{r load}
source("R/process.R")
traces <- readRDS("data/traces.rds")
n_fruit <- length(unique(traces$fruit))
n_trial <- length(unique(traces$trial))
sprintf("%d fruits across %d trials, %d samples per trace.",
        n_fruit, n_trial, sum(traces$fruit == 1))
```

## Example traces with detected minima

Vertical bars mark the detected minimum on each channel
(orange = CH1, green = CH2).

```{r example-trace, fig.height = 3.4}
plot_trace(traces, fruit_id = 4)
plot_trace(traces, fruit_id = 60)
plot_trace(traces, fruit_id = 95)
```

A larger PNG gallery is rendered by `R/batch_plots.R` into
`figures/traces/`.

## Per-fruit summary

```{r summarise}
summary_df <- summarise_traces(traces)
head(summary_df)
```

```{r summary-wide}
summary_wide <- tidyr::pivot_wider(summary_df,
                                   id_cols = c(trial, trial_no, fruit),
                                   names_from = channel,
                                   values_from = t_min_ms)
```

## CH1 vs CH2 scatter (pattern detection)

Each point is one fruit. Most points sit near the CH1 = CH2 diagonal;
vertical excursions are CH2-only outliers (skin defect on the calyx
end of the fruit, in the original interpretation).

```{r scatter, fig.width = 7, fig.height = 5.2}
library(ggplot2)
ggplot(summary_wide, aes(CH1, CH2, colour = trial)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              colour = "grey60") +
  geom_point(size = 2.2, alpha = 0.85) +
  scale_x_continuous(limits = c(0.4, 1.05)) +
  scale_y_continuous(limits = c(0.4, 1.20)) +
  labs(title = "CH2 vs CH1 (synthetic, 2024-2025)",
       x = "CH1 t_min (ms)",
       y = "CH2 t_min (ms)",
       colour = "Trial") +
  theme_minimal(base_size = 12) +
  theme(legend.key.height = unit(0.8, "lines"))

ggsave("figures/scatter_ch1_ch2.png", width = 7, height = 5.2, dpi = 150)
```

## SS1 — CH2 vs CH1 with continuous t_min gradient

Same scatter as above but coloured by the **mean t_min** across the
two channels — a purely geometric ranking that highlights the
diagonal trend without claiming any biophysical firmness model.
Earlier minima cluster in the lower-left (purple/blue); later minima
in the upper-right (orange/red).

```{r ss1-firmness, fig.width = 7.2, fig.height = 5.4}
summary_wide$t_min_mean <- (summary_wide$CH1 + summary_wide$CH2) / 2

ss1_palette <- c("blueviolet", "deepskyblue", "forestgreen",
                 "yellowgreen", "gold", "orange", "red")

ggplot(summary_wide, aes(CH1, CH2, colour = t_min_mean)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              colour = "grey60") +
  geom_point(size = 2.4, alpha = 0.9) +
  scale_x_continuous(limits = c(0.4, 1.05)) +
  scale_y_continuous(limits = c(0.4, 1.20)) +
  scale_colour_gradientn(colours = ss1_palette) +
  labs(title = "SS1 — CH2 vs CH1, coloured by mean t_min",
       x = "CH1 t_min (ms)",
       y = "CH2 t_min (ms)",
       colour = "mean t_min (ms)") +
  theme_minimal(base_size = 12) +
  theme(legend.key.height = unit(1.2, "lines"))

ggsave("figures/ss1_firmness.png", width = 7.2, height = 5.4, dpi = 150)
```

Off-diagonal points indicate asymmetry between stem-end and calyx-end
response on the same fruit.

## Disclaimer

Traces are simulated with a damped Gabor wavelet plus a small initial
transient and Gaussian noise. The repo demonstrates **the processing
pipeline**, not a biological finding.

```{r session}
sessionInfo()
```
